---
title: "Start codon context for Saccharomyces cerevisiae"
author: Edward Wallace, ewjwallace@gmail.com
date: 16 Aug 2018
output:
  # html_notebook: default
  html_document:
    toc: true
    toc_depth: 4
---

# Summary

This presents the renewed analysis of Saccharomyces start codon usage and context. This uses the S. cerevisiae S288C expression and translation estimates from David Weinberg et al 2017.

### Load Packages

```{r setup,message=FALSE,echo=FALSE,cache=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=FALSE,cache=TRUE,
                      results="hide",
                      fig.path="figure/SacCerATGContext-",
                      cache.path="cache/SacCerATGContext-")
library(tidyverse)
library(cowplot)
library(ggrepel)
library(ggseqlogo)
library(Biostrings)

theme_set(theme_cowplot(font_size=12) + 
              theme(strip.background = element_blank()) )



geom_diagline <- function(linetype='solid',size=0.1,colour="grey20",...) {
    geom_abline(slope=1,intercept=0,linetype=linetype,colour=colour)
}

scientific_10 <- function(x) {
    xout <- gsub("1e", "10^{", format(x),fixed=TRUE)
    xout <- gsub("{-0", "{-", xout,fixed=TRUE)
    xout <- gsub("{+", "{", xout,fixed=TRUE)
    xout <- gsub("{0", "{", xout,fixed=TRUE)
    xout <- paste(xout,"}",sep="")
    return(parse(text=xout))
}

scale_x_log10nice <- function(name=waiver(),omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_x_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}

scale_y_log10nice <- function(name=waiver(),omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_y_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}

scale_loglog <- function(...) {
    list(scale_x_log10nice(...),scale_y_log10nice(...))
}

round3 <- function(x) round(x,digits=3)

logmean <- function(x) {
    lx <- log(x)
    exp( mean( lx[is.finite(lx)] ) )
}
# setwd("~/Repos/CryptoTranscriptome2018/ATGContextOthers/")
```


# Here we go

## Expression: RNA abundance and ribosome-protected-fragments

### Load expression data

```{r load_ribosum_Sc,results="show"}

ribosum_Sc <- read_tsv("data_in/ExpressionTranslation/Weinberg2016_SacCer_riboprofiling.txt") %>%
    rename(Gene_name="Gene",Ribo="RPF")

ribosum_Sc %>%
    arrange(desc(RPF)) 

hiTrans_Sc <- ribosum_Sc %>%
    arrange(desc(RPF)) %>%
    head(n=300) %>%
    .$Gene

enoughRNA_Sc <- ribosum_Sc %>%
    arrange(desc(RNA)) %>%
    head(n=3000) %>%
    .$Gene
```

We also calculated hiTrans_Sc, the top 5% (330) translated genes by RPF TPM.

### Ribosome occupancy mostly tracks RNA abundance

```{r plot_TE_Sc,dependson="load_ribosum_Sc",fig.height=3.9,fig.width=4.1}
TE_unusual_Sc <- 
    ribosum_Sc %>%
    filter( (RPF > 1.8e4) |
                (RPF > 100 & RPF > RNA * 5 ) | 
                (RNA > 150 & RPF < RNA / 10 ))


ggplot(data=ribosum_Sc,aes(x=RNA,y=RPF)) +
    geom_diagline() +
    geom_point(size=.8,colour="grey50") +
    geom_point(data=TE_unusual_Sc,aes(colour=Gene),size=1) + 
    geom_text_repel(data=TE_unusual_Sc,aes(label=Gene,colour=Gene)) +
    scale_loglog(limits=c(1e-1,2e4),
                  oob=scales::squish,
                  expand=c(0.01,0.01)) +
    labs(x="RNA abundance (TPM of aORF)",
         y="Ribosome occupancy (TPM of aORF)") +
    theme(legend.position="none")
```

## ATG Context

### Load context data

```{r load_context_Sc,results="show"}
read_ATGcontext_table <- function(file) {
    read_tsv(file,comment="#",skip=1,
             col_names=c("Gene","aATG.context","aATG.pos",
                         "d1.context","d1.posTSS","d1.posATG","d1.frame",
                         "d2.context","d2.posTSS","d2.posATG","d2.frame",
                         "u1.context","u1.posTSS","u1.posATG","u1.frame",
                         "u2.context","u2.posTSS","u2.posATG","u2.frame")
    ) %>% 
        filter(!is.na(aATG.context) ) %>%
        mutate(aATG.context=toupper(aATG.context))
}
ctbl_big_Sc <- 
    read_ATGcontext_table("data_in/SacCer_allATGcontext.table.txt")

ctbl_big_Sc
```

### Annotated ATGs have a Kozak consensus sequence

```{r aATG_context_Sc,dependson="load_context_Sc",fig.width=4,fig.height=1}
ggseqlogo(data=ctbl_big_Sc$aATG.context,namespace="TCAG") + 
    theme_nothing()
```

### Highly translated Annotated ATGs have a Kozak consensus sequence

That's for hiTrans_Sc, the top 5% (330) translated genes by RPF TPM.

```{r a_hi_context_Sc,dependson=c("load_context_Sc","load_ribosum_Sc"),fig.width=4,fig.height=1}
ggseqlogo(data=filter(ctbl_big_Sc,Gene %in% hiTrans_Sc)$aATG.context,
                      namespace="TCAG") + 
    theme_nothing()
```

### Cytoplasmic Ribosome Annotated ATGs have a Kozak consensus sequence

```{r cytoribo_context_Sc,dependson="load_context_Sc",fig.width=4,fig.height=1}
cytoribolist_Sc <- read_tsv("data_out/ComplexOrthologs/cytoribosome_PTRcurated4.txt",
                        comment="#") %>%
    filter(SPECIES=="YEAST") %>%
    .$GENE_ID
ggseqlogo(data=filter(ctbl_big_Sc,Gene %in% cytoribolist_Sc)$aATG.context,
                      namespace="TCAG") + 
    theme_nothing()
```



### Downstream ATGs don't have a consensus

First downstream ATG

```{r d1_context_Sc,dependson="load_context_Sc",fig.width=4,fig.height=1}
ggseqlogo(data=filter(ctbl_big_Sc,nchar(d1.context) ==24)$d1.context,
                      namespace="TCAG") + 
    theme_nothing()
```

### Downstream ATGs in frame and highly translated don't have a consensus

Except for 3rd-codon-position bias.

```{r d1_hi_frame0_context_Sc,dependson=c("load_context_Sc","load_ribosum_Sc"),fig.width=4,fig.height=1}
ggseqlogo(data=filter(ctbl_big_Sc,
                      nchar(d1.context) == 24,
                      d1.frame==0,
                      Gene %in% hiTrans_Sc)$d1.context,
                      namespace="TCAG") + 
    theme_nothing()
```

## Calculate Information content and scores of consensus motif

### Calculate a wide and a narrow consensus sequence

Calculate motif score against the position weight matrix (pwm) for both **narrow** (-5 from ATG through to ATG) and **wide** (-9 from ATG to +3) kozak consensus motif. These motifs are taken from the top 5% highly translated genes.

```{r kozak_consensus_Sc,dependson=c("load_context_Sc","load_ribosum_Sc"),fig.width=4,fig.height=1}
kozak_w_PFM_Sc <- filter(ctbl_big_Sc,Gene %in% hiTrans_Sc) %>%
    .$aATG.context %>%
    str_sub(start=4L,end=18L) %>%
    str_to_upper() %>%
    consensusMatrix() 

kozak_w_PWM_Sc <- PWM(kozak_w_PFM_Sc)
kozak_n_PWM_Sc <- kozak_w_PFM_Sc[,5:12] %>% PWM()
```


```{r write_PFM_Sc,dependson="kozak_consensus_Sc",results="show"}
cat(
    "# PFM_hiTrans_Sc.txt
# Position Frequency Matrix for aATGs of 330 highest-translated genes from Sc
# This is of the wide ATG context NNNNNNNNNATGNNN
# Column is position, Row is nt count in order: ACGT.
#
# Edward Wallace, June 2018
",
file="data_out/PFM_hiTrans_Sc.txt")
kozak_w_PFM_Sc %>%
    as.tibble() %>%
    write_tsv("data_out/PFM_hiTrans_Sc.txt",
              append = TRUE,col_names=FALSE)
```

### Estimate the information content 

Using the sequence logo, details on https://en.wikipedia.org/wiki/Sequence_logo

```{r kozak_inf_Sc,dependson=c("kozak_consensus_Sc","cytoribo_context_Sc"),fig.width=4,fig.height=1,results="show"}
infon1 <- function(counts,ssize=4,pseudocount=0.5) {
    # calculate information content for one position
    # as described in https://en.wikipedia.org/wiki/Sequence_logo
    # counts: position frequency matrix
    # ssize: alphabet size
    # pseudocount: pseudocount to add to remove zeroes
    
    counts <- counts + pseudocount # add pseudocount
    ntot = sum(counts)[1] # total number of sequences
    freqs = counts / ntot # freqency
    Hentropy = - sum( freqs * log2(freqs) ) # Shannon entropy 
    ecorrection = ( ssize - 1 ) / ( log(2) *  ntot ) # small-sample correction
    information = log2(ssize) - Hentropy # information content
    return(information)
}

infonall <- function(pfm,ssize=4,pseudocount=0.5) {
    apply(pfm,2,infon1)
}

infontot <- function(pfm,ssize=4,pseudocount=0.5) {
    apply(pfm,2,infon1) %>% sum()
}

infonfromseqs <- function(seqs,sstart=1L,send=-1L,ssize=4,pseudocount=0.5) {
    seqs %>% 
    str_sub(start=sstart,end=send) %>%
    str_to_upper %>%
    consensusMatrix %>%
    infontot(ssize=ssize,pseudocount=pseudocount)
}


inf_content_narrow <- 
    tribble(
        ~Genes, ~ATG, ~Width, ~Infon,
        "All", "aATG", "narrow", 
        ctbl_big_Sc %>%
            .$aATG.context %>%
            infonfromseqs(sstart=8L,send=12L),
        "HiTrans", "aATG", "narrow", 
        ctbl_big_Sc %>%
            filter(Gene %in% hiTrans_Sc) %>% 
            .$aATG.context %>%
            infonfromseqs(sstart=8L,send=12L),
        "CytoRibo", "aATG", "narrow", 
        ctbl_big_Sc %>%
            filter(Gene %in% cytoribolist_Sc) %>% 
            .$aATG.context %>%
            infonfromseqs(sstart=8L,send=12L),
        "All", "d1ATG", "narrow", 
        ctbl_big_Sc %>%
            .$d1.context %>%
            na.omit() %>%
            infonfromseqs(sstart=8L,send=12L),
        "HiTrans", "d1ATG", "narrow", 
        ctbl_big_Sc %>%
            filter(Gene %in% hiTrans_Sc) %>% 
            .$d1.context %>%
            na.omit() %>%
            infonfromseqs(sstart=8L,send=12L),
        "CytoRibo", "d1ATG", "narrow", 
        ctbl_big_Sc %>%
            filter(Gene %in% cytoribolist_Sc) %>% 
            .$d1.context %>%
            na.omit() %>%
            infonfromseqs(sstart=8L,send=12L)
    )

write_tsv(inf_content_narrow %>%
              mutate_if(is.numeric,round,digits=3),
          "inf_content_narrow_Sc.txt")
inf_content_narrow
```

Information content in bits of highly-translated consensus (excluding 6 bits from ATG), **narrow** is `r infontot(kozak_w_PFM_Sc[,5:9]) %>% round(2)`, of **wide** is `r infontot(kozak_w_PFM_Sc[,-(10:12)]) %>% round(2)`.


### Calculate scores of aATG, dATG, uATG against Kozak consensus

```{r calc_scores_Sc, dependson="kozak_consensus_Sc"}

PWMscore <- function(seqs,pwm,startl) {
    # calculate scores for a character vector seqs against
    # pwm from given starting position
    # after checking strings are non-missing and start with ACTG
    vapply(str_sub(seqs,start=startl),
           function(ss) {
               if ( is.na(ss) ) {
                   return(NA)
               } else if ( str_detect(ss,"\\A[ACTG]") ) {
                   return( PWMscoreStartingAt(pwm,ss) )
               } else {
                   return(NA) 
               } 
           } ,
           FUN.VALUE = 0)
}

PWMscoren <- function(seqs,pwm=kozak_n_PWM_Sc,startl=8) {
    PWMscore(seqs,pwm,startl) 
}

PWMscorew <- function(seqs,pwm=kozak_w_PWM_Sc,startl=4) {
    PWMscore(seqs,pwm,startl) 
}

is.context <- function(x) str_detect(x,"context")

scores_Sc <- 
    ctbl_big_Sc %>%
    transmute(Gene,
              aATG.scorekn = PWMscoren(str_to_upper(aATG.context)),
              d1.scorekn   = PWMscoren(d1.context),
              u1.scorekn   = PWMscoren(u1.context),
              # aATG.scorekw = PWMscorew(str_to_upper(aATG.context)),
              # d1.scorekw   = PWMscorew(d1.context),
              # u1.scorekw   = PWMscorew(u1.context),
              NULL
    ) %>%
    mutate(d1vsan = d1.scorekn - aATG.scorekn,
           u1vsan = u1.scorekn - aATG.scorekn,
           # d1vsaw = d1.scorekw - aATG.scorekw,
           # u1vsaw = u1.scorekw - aATG.scorekw,
           NULL) 

ribotbl_scoresenough_Sc <-
    ribosum_Sc %>%
    filter(Gene %in% enoughRNA_Sc) %>%
    # left_join(ribotbl_Sc,by="Gene") %>% 
    left_join(scores_Sc,by="Gene")

```

We calculate scores using Biostrings::PWMscoreStartingAt.

The best description I could find of this method is: https://support.bioconductor.org/p/61520/

It is just the sum of the matrix product of the PWM with the sequence.


Write scores to file *scores_kozak_Sc.txt*.

```{r write_scores_Sc, dependson="calc_scores_Sc",results="show"}
scores_Sc
cat(
    "# scores_kozak_Sc.txt
# Scores of ATG context against Kozak sequence, C. neoformans serotype D Sc
# Kozak sequence derived from 330 highest-translated genes by RPF
#
# Gene: Sc systematic gene name
# aATG.scorekn: score of annotated ATG against narrow Kozak sequence (NNNNNATG)
# d1.scorekn: score of 1st downstream ATG agains narrow
# xx.scorekw: analogous scores against wide Kozak sequence (NNNNNNNNNATGNNN)
#
# Edward Wallace, June 2018
",
file="data_out/scores_kozak_Sc.txt")
scores_Sc %>%
    mutate_if(is.numeric,round3) %>%
    write_tsv("data_out/scores_kozak_Sc.txt",
              append = TRUE,col_names=TRUE)
```

### Plot against narrow consensus (-5 to ATG)

```{r plot_scores_n_Sc, dependson="calc_scores_Sc", fig.width=5,fig.height=1.8}

sbandwidthn = 0.025

ggplot(data=scores_Sc) +
    geom_density(aes(x=aATG.scorekn,colour="aATG"),
                 kernel="rectangular",bw=sbandwidthn) +
    geom_density(aes(x=d1.scorekn,colour="dATG"),
                 kernel="rectangular",bw=sbandwidthn) +
    geom_density(aes(x=u1.scorekn,colour="uATG"),
                kernel="rectangular",bw=sbandwidthn) +
    geom_density(data=filter(scores_Sc,Gene %in% hiTrans_Sc),
                 aes(x=aATG.scorekn,colour="aATG, top 5%"),
                 kernel="rectangular",bw=sbandwidthn) +
    scale_colour_manual(values=c("aATG"="dodgerblue",
                                 "aATG, top 5%"="darkblue",
                                 "uATG"="red",
                                 "dATG"="green3") ) +
    labs(x="ATG Kozak score, Sc") +
    theme(axis.line.y=element_blank(),axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),axis.title.y=element_blank(),
          legend.title=element_blank())
```

## Compare aATG and dATG context by gene

### Most dATG scores are less than aATG scores

```{r dvsaATG_scoresdens_Sc, dependson="calc_scores_Sc", fig.width=3.5,fig.height=1.8}

ggplot(data=scores_Sc) +
    geom_vline(xintercept=0,size=0.3) +
    geom_density(aes(x=d1vsan,colour="narrow"),kernel="rectangular") +
    # geom_density(aes(x=d1vsaw,colour="wide"),kernel="rectangular")  +
    labs(x="difference in score, dATG vs aATG",
         colour="context") +
    theme(axis.line.y=element_blank(),axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),axis.title.y=element_blank(),
          legend.position=c(0.8,0.8))
```

### For highly translated genes, most dATG scores are much less than aATG

```{r dvsaATG_scoresn_Sc, dependson="calc_scores_Sc", fig.width=3.5,fig.height=3.5}
highdiffn_dvsaATG_Sc <- scores_Sc %>%
    arrange(desc(d1vsan)) %>%
    head(n=330) %>%
    select(Gene,aATG.scorekn,d1.scorekn)

ggplot(data=scores_Sc %>% filter(!is.na(d1.scorekn)),
       aes(x=aATG.scorekn,y=d1.scorekn)) +
    geom_diagline() +
    geom_point(size=0.5,colour="grey50") +
    geom_point(data=highdiffn_dvsaATG_Sc,
               colour="red",size=1) +
    geom_point(data=scores_Sc %>% 
                   filter(Gene %in% hiTrans_Sc),
               colour="blue",size=1) +
    geom_point(data=highdiffn_dvsaATG_Sc %>% 
                   filter(Gene %in% hiTrans_Sc),
               colour="purple4",size=1.3) +
    labs(x="aATG score, narrow",
         y="dATG score, narrow")
```

Red: high dATG vs aATG **Kozak** score.
Blue: highly translated.
Purple: both.

## Genes with unusual dATG vs aATG narrow score

Those genes are in this list:

```{r highdiffn_dvsaATG_Sc,dependson="dvsaATG_scoresn_Sc",results="show"}
highdiffn_dvsaATG_Sc
```

* SRL2 / YLR082C unknown function, localized to nucleus and cytoplasm
* HSF1 / YGL073W heat shock transcription factor 1, known to shuttle between nucleus and cytoplasm, weird reported mitochondrial localization
* YER156C, unknown function,  localizes to the nucleus and cytoplasm in a large-scale study
* INP1 / YMR204C, peripheral membrane protein of peroxisomes
* CIN8 / YEL061C, kinesin motor protein involved in mitosis. Reported mitochondrial localization.
* ERI1 / YPL096C-A, ER-associated Ras Inhibitor
* GRX2 / YDR513W, Cytoplasmic/mitochondrial glutaredoxin, predicted mito localization sequence before 2nd in-frame ATG
* YNR025C, Dubious open reading frame
* PHO86 / YJL117W, ER-resident protein involved in packaging phosphate transporter Pho84 to vesicles
* TPK1 / YJL164C, cAMP-dependent protein kinase catalytic subunit, localizes to nucleus and cytoplasm

These genes are also highly translated:

```{r highdiffn_dvsaATG_hiTrans_Sc,dependson="dvsaATG_scoresn_Sc",results="show"}
highdiffn_dvsaATG_Sc %>% 
                   filter(Gene %in% hiTrans_Sc) 
```

* HNT1 / YDL125C, Adenosine monophosphoramidase involved in nucleotide metabolism, localizes to nucleus and cytoplasm
* SUI1 / YNL244C eIF1, enforces accurate ATG selection and has conserved poor start codon context.
* RPL39 / YJL189W, Ribosomal protein L39

There is no clear pattern in these.

## Compare score difference to localization predictions

We need to put the corresponding SacCer proteome through mitofates in order to run these analyses.

### Load predictions from mitofates

In input file *SacCer_mitofates_parsed.table.txt*.

```{r load_mitofates_Sc}
mitofates_Sc <- read_tsv("data_in/SacCer_mitofates_parsed.table.txt",
                            comment="#")
mitocounts_Sc <- 
    mitofates_Sc %>% 
    group_by(Pred_preseq) %>% 
    count() 
mitobreaks_Sc <- mitocounts_Sc %>%
    mutate(label=paste0(Pred_preseq," (n = ",n,")")) %>%
    .$label %>%
    set_names(mitocounts_Sc$Pred_preseq)

mitopredlist_Sc <- mitofates_Sc %>%
    filter(Pred_preseq =="Yes") %>%
    .$Gene
```


### SacCer Genes with high dATG vs aATG score are not enriched in mitochondrial presequences

```{r dvsaATG_scoresdiffn_mitofates_Sc, dependson=c("calc_scores_Sc","load_mitofates_Sc"), fig.width=5,fig.height=1.8}

ggplot(data=scores_Sc %>% 
           filter(!is.na(d1.scorekn)) %>%
           left_join( mitofates_Sc %>% 
                          select(Gene,Prob_preseq,Pred_preseq)
                      ) ,
       aes(x=d1.scorekn-aATG.scorekn,colour=Pred_preseq)) +
    geom_vline(xintercept=0,size=0.3) +
    geom_density(kernel="rectangular",bw=0.025) +
    scale_colour_manual(values=c("Yes"="magenta","No"="grey50"),
                        labels=mitobreaks_Sc)  + 
    labs(x="Kozak score, difference dATG - aATG", colour="Mito. pred") +
    theme(axis.line.y=element_blank(),axis.ticks.y=element_blank(),
          axis.text.y=element_blank(),axis.title.y=element_blank())
```

### SacCer mito-localized genes also do not have a distinctive aATG context

Although the +5 T is striking.

```{r aATG_motif_mitofates_Sc, dependson=c("calc_scores_Sc","load_mitofates_Sc"), fig.width=4,fig.height=1}

ggseqlogo(data=ctbl_big_Sc %>%
              filter(Gene %in% mitopredlist_Sc) %>%
              .$aATG.context,
          namespace="TCAG")  + 
    theme_nothing()

```


<a href="#top">Back to table of contents</a>
